### 2009-12-09 docs/marcello-dissertation/thesis/chapter4-netbeams-in-depth-requirements-analysis-and-specification.tex
Update of chapter 5 after Niki's reviews and my verification using MS Word.
git-svn-id: svn://localhost/netbeams/branches/marcello/persistence@688 559edde2-a4d4-44e0-96fc-4035e40d7396

As discussed in chapter 2, different properties of a sensor network and the
nature of the collected data must be taken into account in order to provide a
data persistence layer for a given sensor netwokr. Similarly, in order to
better assist one analysis of such functionality, chapter 3 proposed a set of
data persistence taxonomies related to different properties of the collected
data life cycle. In this way, the goal of this chapter is to describe the
fundamental properties of the case study used for this work, that is, NetBEAMS,
in order to ``set the stage'' for the selection of a database technology for it.
In this way, given the fact that NetBEAMS provides an automated infrastructure
solution for SF-BEAMS, this last component will be covered first. Then, this
work proposes a categorization of NetBEAMS to the face of the taxonomies, and
finally discusses the discusses the requirements for a data persistence.

@@

As described by \cite{netbeams2009}, NetBEAMS is a joint venture between the
department of Biology and Computer Science at San Francisco State University,
whose goal was to automate the operational execution of SF-BEAMS. On the other
hand, the San Francisco Bay Environmental Assessment and Monitoring Station, or
SFBEAMS \cite{sfbeams2006}, is an instance of an environmental sensor
network whose primary focus is the study of complex marine and estuarine
environments using the SF-BEAMS sensors. The network is deployed offshore of
its pier located on the banks of the San Francisco Bay at the Tiburon Island,
California, being operated by the department of Biology from San Francisco
State University. In this way, NetBEAMS is an experimental component that
utilizes the SF-BEAMS infrastructure to operate. This section details SF-BEAMS
in general, and presents the requirements for data persistence.

@@

In general, the SF-BEAMS network infrastructure contains varying a number of
wired and wireless devices standing on pylons docked at the pier. Each of them
is responsible for observing different conditions of the area, having its own
mechanisms for internal storage for the collected data. In this way, data can
be directly transferred to the labs via Ethernet cables or collected manually
with a laptop computer. The current data collection process for SF-BEAMS is
described in figure
\ref{fig:SF-BEAMS-system-architecture}\cite{sfbeams-current-system}. Upon
collecting data from sensors, the RTC staff uses automation scripts written in
Matlab \cite{matlab} to process, index and distribute the raw data in
different formats. One of these formats is the OPeNDAP \cite{opendap}, which
is widely used at research institutions that promote easier data exchange
among them. The SF-BEAMS website provides access to the collected data through
the Internet at http://sfbeams.sfsu.edu:8080/opendap. An example of an HTTP
Request from accessing the data for a specific YSI is shown in Listing
\ref{file:rtc-ysi-opendap}.

@@

As described in section \ref{sec:sn-infrastructure}, sensor devices produce
its observed data based on properties of measurements defined by its
manufacturer. NetBEAMS has used ``YSI 6600 ESD V2", as seen in Figure
\ref{fig:ysi-device}, one of the sensors devices used by SF-BEAMS. It is a
powerful water quality monitoring device, capable of producing around 52 bytes
on a single real-time data stream reading, as shown on Table
\ref{tab:ysi-data-stream}.

\begin{figure}[!b]
  \centering
  \includegraphics[scale=0.7]{../diagrams/ysi-device}
  \caption{Picture of the case study sonde device: YSI 6600 ESD V2}
  \label{fig:ysi-device}
\end{figure}

@@

In order to have access to the observed data, the network staff downloads the
data using one of the device's connection such as the RS-232 serial connector
\cite{rs232}. Then, the RTC labs receives the collected data transferred from
the sensor devices to be used for archival purposes. According to one of its
staff members, the infrastructure of the SF-BEAMS in was comprised of
the following components (as of May 2009):

\begin{itemize}
  \item 5 YSI Sonde devices, among others, in operation at the RTC pier;
  \item The sampling frequency rate is configured in ranges of either 1, 6 or
  15 minutes, depending on the specification;
  \item The processed data, used for distribution over the Internet, is
   contains information regarding the time of the data collection.
\end{itemize}

Considering the infrastructure aforementioned, the data load in the server-side
of the RTC's laboratories produced by the YSI sonde sensors devices can be
estimated by analyzing the data distribution over a period of one year, as
shown in Table \ref{tab:ysi-data-distribution}:

\begin{table}[!b]
    \label{tab:ysi-data-distribution}
    \caption{Amount of data produced by the RTC's YSI sondes}
        \begin{center}
        \begin{tabular}{|c|c|c|c|c|c|c|}\hline 
        \textbf{YSIs} & \textbf{Rate} & \textbf{Hourly} & \textbf{Daily} &
        \textbf{Weeky} & \textbf{Monthly} & \textbf{Yearly}\\\hline 
        1 & 1 min & 3.04 Kb & 73.12 Kb & 511.87 Kb & 1.99 Mb & 23.99 Mb\\\hline 
        5 & 6 min & 15.23 Kb & 365.62 Kb & 2.5 Mb & 9.99 Mb & 119.97 Mb\\\hline 
        1 & 15 min & 0.5 Kb & 12.18 Kb & 85.31 Kb & 341.25 Kb & 3.99 Mb\\\hline 
        5 & 1 min & 2.54 Kb & 60.93 Kb & 426.56 Kb & 1.67 Mb & 19.99 Mb\\\hline
        1 & 6 min & 0.2 Kb & 4.87 Kb & 34.12 Kb & 136.5 Kb & 1.6 Mb\\\hline 
        5 & 15 min & 1.0 Kb & 24.37 Kb & 170.62 Kb & 682.5 Kb & 7.99 Mb\\\hline
        \end{tabular}
        \end{center}
\end{table}

By analyzing the maximum amount of produced data by the YSI sonde device is
around hundreds of Megabytes a year, as \textbf{483,840} samples are generated
by one single device.

@@

\begin{itemize}
  \item \textbf{The Purpose of Sensor Data}: the data from SF-BEAMS is solely
  used for the purpose of Data Archival, since they are stored in the RTC's
  lab;
  \item \textbf{The Location of the Sensor Data}: it is clear that the purpose
  of data is directly related to its location and, as a consequence, the
  strategy of \textbf{External Data Storage} is used to store its collected
  data at the network sink at the RTC's labs;
  \item \textbf{Data Model}: since the format uses by the RTC staff to share
  data among researcher is in the OPEnDAP format, it is clear that the data
  model used is the tabular data model, since it uses the comma-delimited
  files. In this way, the \textbf{Schema-less} is used;
  \item \textbf{Data Description}: once the collected data reaches the RTC
  lab, a quality assurance process is executed, and the data is transformed
  into the format used by OPEnDAP. This conversion carries data regarding
  the time of the data collection, and therefore, the use of one of the
  \textbf{Time Dimensions} such as the valid time. Since the files are stored
  with specific file name structure, using timestamps to describe the data, the
  use of transaction time may be also considered. Finally, the contents of the
  files contain the \textbf{Data Identity} as shown in the samples of
  the collected in Listing \ref{file:rtc-ysi-opendap};
  \item \textbf{Query Processing Mechanism}: similar to the data stored in a
  centralized way, the query processing is also \textbf{Centralized}, as the
  data is stored on file system;
  \item \textbf{Data Volume}: the volume of data produced is in the range of
  small. It is related to the volume of data produced by the sensor devices,
  which is numerical data streams, and the total number of sensor devices;
  \item \textbf{System Organization}: SF-BEAMS runs on a Single System, as
  the collected data is stored directly in the file-system after previous 
  processing. However, they provide the collected data to the Internet using
  the OPeNDAP Hyrax system as a middleware.
\end{itemize}

@@

Although the execution of the RTC's sensor network can be operated as it
described in the previous sections, \cite{netbeams2009} described some of the
operational challenges faced by the RTC staff during regular activities of
data collection process. It was clear to the research group that SF-BEAMS
could have not only its data gathering process automated, but also its data
management and distribution. By the use of COTS\footnote{Common-Off-The-Shelf
designates a product that is produced and sold in bulk} embedded devices and
open-source \cite{open-source} software, the research group developed a second
version of NetBEAMS, a component-based approach suggested improving operational
activities from SF-BEAMS using systems automation. In brief, one of the open
problems of NetBEAMS was regarding Data Persistence, the primary motivation of
this dissertation.

The Networked Bay Environmental Assessment and Monitoring System, or Net-BEAMS,
offers the Data Sensor Platform (DSP) \cite{netbeams2009} as the system
architecture that can address the operation of SF-BEAMS. In this way, in-depth
documentation describing the DSP Platform, including its architecture and data
gathering process, is provided startng in the appendix section
\ref{sec:dsp-details}.

The scope of this work is defined as to provide a persistence layer for
NetBEAMS, making a selection of a persistence layer that is an exception to the
taxonomical classification of SF-BEAMS. Furthermore, the technology to be
selected for the persistence layer must take into account the requirements of
NetBEAMS primary users such as Computer Science researchers, as well as end
users such as Biologists from the RTC laboratories. The requirements for the
data persistece specifically for NetBEAMS is described in the following section.

@@

The main reason for the operation of SF-BEAMS using NetBEAMS's automated
approach is that it can potentially reduce operational costs and improve the
data collection process. In review of this, the scope of a persistence layer
can be summarized as a group of functional and non-functional requirements in
order to select a database system for the sensor network supported by NetBEAMS.

@@

\begin{itemize}
  \item The data model used to describe the data must not impose restrictions
  to the users of the system: Biologists and students without expertise in
  Database Systems;
  \item Data Representation must be similar to the ones used by RTC, with a
  more human-readable format;
  \item The system must be scalable with a low degree of maintenance, in a
  way that the data gathering process does not impose less service interruption;
  \item Data must be searchable in near-real-time with outstanding performance;
  \item The solution must also give support to export capabilities, which can
  be possibly used to match the RCT's requirements of the OPeNDAP formatt;
  \item The database system must be free of charge, following the
  implementation specifications of NetBEAMS for Open-source software.
\end{itemize}

After presenting NetBEAMS and its infrastructure, the the next step is the
actual analysis of existing database systems that fits to the requirements
and characteristics of NetBEAMS. Meanwhile, more details regarding the
architecture of NetBEAMS, as well as development guidelines, can be reached in
the appendix of chapter 10.

>>>>

As discussed in Chapter 2, different properties of a sensor network and the
nature of the collected data must be taken into account in order to provide a
data persistence layer for a given sensor network. Similarly, in order to
better assist one's analysis of such functionality, Chapter 3 proposed a set of
data persistence taxonomies related to different properties of the collected
data life cycle. The goal of this chapter is to describe the fundamental properties 
of the case study used for this work, that is, NetBEAMS, in order to lay the 
groundwork as it relates to the selection of the database technology. NetBEAMS 
provides an automated infrastructure solution for SF-BEAMS, and this last component 
will be covered first. Then, this work proposes a categorization of NetBEAMS to 
the face of the taxonomies.

@@

As described by \cite{netbeams2009}, NetBEAMS is a joint venture between the
department of Biology and Computer Science at San Francisco State University,
whose goal is to automate the operational execution of SF-BEAMS. The San Francisco
Bay Environmental Assessment and Monitoring Station, or SFBEAMS \cite{sfbeams2006}, 
is an environmental sensor network system whose primary focus is the study of complex 
marine and estuarine environments using the SF-BEAMS sensors. The sensors are deployed off a
pier located on the San Francisco Bay, in Tiburon, California; and is operated by 
the Romberg Tiburon Center for Environmental Studies (RTC) a part of San Francisco
State University. NetBEAMS is an experimental component that utilizes the SF-BEAMS 
infrastructure to operate. This section details SF-BEAMS in general, and presents 
the requirements for data persistence.

@@

In general, the SF-BEAMS network infrastructure contains a varying number of
wired and wireless devices attached to pylons at the pier. Each of them
is responsible for observing different conditions of the area, having its own
mechanisms for internal storage for the collected data. In this way, data can
be directly transferred to the labs via Ethernet cables or collected manually
with a laptop computer. The current data collection process for SF-BEAMS is
described in Figure
\ref{fig:SF-BEAMS-system-architecture}\cite{sfbeams-current-system}. Upon
collecting data from sensors, the RTC staff uses automation scripts written in
Matlab \cite{matlab} to process, index and distribute the raw data in
different formats. One of these formats is the OPeNDAP \cite{opendap}, which
is widely used at research institutions to promote easier data exchange
amongst them. The SF-BEAMS website provides access to the collected data through
the Internet at http://sfbeams.sfsu.edu:8080/opendap. An example of an HTTP
Request from accessing the data for a specific YSI \cite{YSI-Sonde} is shown in
Listing \ref{file:rtc-ysi-opendap}.

@@

As described in section \ref{sec:sn-infrastructure}, sensor devices produce
the observed data based on properties of measurements defined by its
manufacturer. NetBEAMS has used ``YSI 6600 ESD V2" \cite{YSI-Sonde}, as seen in
Figure \ref{fig:ysi-device}, one of the sensors devices used by SF-BEAMS. It is a
powerful water quality-monitoring device, capable of producing around 52 bytes
on a single real-time data stream reading, as shown on Table
\ref{tab:ysi-data-stream}.

\begin{figure}[!b]
  \centering
  \includegraphics[scale=0.7]{../diagrams/ysi-device}
  \caption{Picture of the case study Sonde device: YSI 6600 ESD V2}
  \label{fig:ysi-device}
\end{figure}

@@

In order to have access to the observed data, the network staff downloads the
data using one of the device's connections such as the RS-232 serial connector
\cite{rs232}. Then, they upload the data into their data servers where they are
processed and archived for historical analysis. According to one
staff member, the infrastructure of the SF-BEAMS is comprised of the 
following components (as of May 2009):

\begin{itemize}
  \item 5 YSI Sonde devices, among others, in operation at the RTC pier;
  \item The sampling frequency rate is configured in ranges of either 1, 6 or
  15 minutes, depending on the specification;
  \item The processed data, used for distribution over the Internet,
   contains information regarding the time of the data collection.
\end{itemize}

Considering the infrastructure, the data load in the server-side
of RTC's laboratories produced by the YSI Sonde sensors devices can be
estimated by analyzing the data distribution over a period of one year, as
shown in Table \ref{tab:ysi-data-distribution}:

\begin{table}[!b]
    \label{tab:ysi-data-distribution}
     \begin{center}
      \begin{tabular}{|c|c|c|c|c|c|c|}\hline 
        \textbf{YSIs} & \textbf{Rate} & \textbf{Hourly} & \textbf{Daily} &
        \textbf{Weekly} & \textbf{Monthly} & \textbf{Yearly}\\\hline 
        1 & 1 min & 3.04 Kb & 73.12 Kb & 511.87 Kb & 1.99 Mb & 23.99 Mb\\\hline 
        5 & 6 min & 15.23 Kb & 365.62 Kb & 2.5 Mb & 9.99 Mb & 119.97 Mb\\\hline 
        1 & 15 min & 0.5 Kb & 12.18 Kb & 85.31 Kb & 341.25 Kb & 3.99 Mb\\\hline 
        5 & 1 min & 2.54 Kb & 60.93 Kb & 426.56 Kb & 1.67 Mb & 19.99 Mb\\\hline
        1 & 6 min & 0.2 Kb & 4.87 Kb & 34.12 Kb & 136.5 Kb & 1.6 Mb\\\hline 
        5 & 15 min & 1.0 Kb & 24.37 Kb & 170.62 Kb & 682.5 Kb & 7.99 Mb\\\hline
        \end{tabular}
      \end{center}
    \caption{Amount of data produced by the RTC's YSI sondes}
\end{table}

The maximum amount of data produced by the YSI Sonde device is hundreds of Megabytes
a year, as \textbf{483,840} samples are generated by one single device.

@@

\begin{itemize}
  \item \textbf{The Purpose of Sensor Data}: the data from SF-BEAMS is exclusively
  used for the purpose of Data Archival, since they are stored in RTC's lab;
  \item \textbf{The Location of the Sensor Data}: it is clear that the purpose
  of the data is directly related to the sensors location and, thus, the
  strategy of \textbf{External Data Storage} is used to store its collected
  data at the network sink at the RTC site;
  \item \textbf{Data Model}: since the format used by the RTC staff to share
  data with researchers is in the OPeNDAP format, the data
  model used is the tabular data model, since it uses the comma-delimited
  files. In this way, the \textbf{Schema-less} model is used;
  \item \textbf{Data Description}: once the collected data reaches the RTC
  lab, a quality assurance process is executed, and the data is transformed
  into the format used by OPeNDAP. This conversion carries data regarding
  the time of the data collection, therefore, one of the
  \textbf{Time Dimensions} such as the valid time is used. Since the files are stored
  with specific file name structure, using timestamps to describe the data, the
  use of transaction time may be also considered. Finally, the contents of the
  files contain the \textbf{Data Identity} as shown in the samples of
  the collected in Listing \ref{file:rtc-ysi-opendap};
  \item \textbf{Query Processing Mechanism}: similar to the data stored in a
  centralized way, the query processing is also \textbf{Centralized}, as the
  data is stored on file system;
  \item \textbf{Data Volume}: the volume of data produced is characterized by small.
 It is related to the volume of data produced by the sensor devices,
  which is numerical data streams, and the total number of sensor devices;
  \item \textbf{System Organization}: SF-BEAMS runs on a Single System, as
  the collected data is stored directly in the file-system after previous 
  processing. However, the collected data is provided to the Internet using
  the OPeNDAP Hyrax system as a middleware.
\end{itemize}

@@

Although the execution of RTC's sensor network can be operated as it
was mentioned in the previous sections, \cite{netbeams2009} described some of the
operational challenges faced by the RTC staff during regular activities of the
data collection process. It was clear to the research group that SF-BEAMS
could have not only its data gathering process automated, but also its data
management and distribution. By the use of COTS\footnote{Common-Off-The-Shelf
designates a product that is produced and sold in bulk} embedded devices and
open-source \cite{open-source} software, the research group developed a second
version of NetBEAMS, a component-based approach which suggested the improvement of 
operational activities from SF-BEAMS using systems automation. In brief, one of the 
ongoing problems of NetBEAMS was regarding Data Persistence, the primary motivation of
this report.

The Networked Bay Environmental Assessment and Monitoring System, or Net-BEAMS,
offers the Data Sensor Platform (DSP) \cite{netbeams2009} as the system
architecture that can address the operation of SF-BEAMS. The in-depth
documentation describing the DSP Platform, including its architecture and data
gathering process is provided at the online documentation 
\cite{netbeams-dsp-architecture}.

The scope of this work is defined as to provide a persistence layer for
NetBEAMS, making a selection of a persistence layer that is an exception to the
taxonomical classification of SF-BEAMS. Furthermore, the technology to be
selected for the persistence layer must take into account the requirements of
NetBEAMS primary users such as Computer Science researchers, as well as end
users such as Biologists from the RTC laboratories. The requirements for the
data persistence specifically for NetBEAMS are described in the following
section.

@@

The main reason for the operation of SF-BEAMS using NetBEAMS's automated
approach is that it can potentially reduce operational costs and improve the
data collection process. In view of this, the scope of a persistence layer
can be summarized as a group of functional and non-functional requirements in
order to select a database system for the sensor network supported by NetBEAMS.

@@

\begin{itemize}
  \item The data model used to describe the data must not impose restrictions
  to the users of the system (i.e. Biologists and students without expertise in
  Database Systems);
  \item Data Representation must be similar to those used by RTC, with a
  more human-readable format;
  \item The system must be scalable with a low degree of maintenance, in a
  way that the interruptions to the data gathering process are infrequent;
  \item Data must be searchable in near-real-time with good performance;
  \item The system must also support export capabilities, which may
  be used to match RCT's requirements of the OPeNDAP format;
  \item The database must be free of charge, following the
  implementation specifications of NetBEAMS for Open-source software.
\end{itemize}

In conclusion, after detailing the SF-BEAMS sensor network, and describing the 
infrastructure of NetBEAMS, its the motivation and goals for this project were 
presented. The next chapter is the actual analysis of existing database systems 
mentioned in Chapter 2, which aligns with the requirements and characteristics of 
NetBEAMS. Meanwhile, more details regarding the architecture of NetBEAMS, as well as
the development guidelines can be found in the appendix of chapter 10.
